Learning \strips\ action models with classical planning

### 2018-02-23 doc/aij18/aij18.tex
added updates to the AIJ18 paper

A first description of the compilation and its extensions previously appeared in several conference papers~\cite{aineto2018learning}. Compared to the conference papers, the present paper includes the following novel material:
\begin{itemize}
\item Unifies the formulation of our approach for learning \strips\ action models from state observations and from labeled plans.
\item Shows that our compilation is also valid for the particular scenario where the observed states are not full states but {\em partial states}, in the sense that some of the observed fluents (either with positive or negative value) are missing.
\item Proposes a distance metric to assess how well a \strips\ action model matches a given set labeled plans. Defines the compilation to compute this distance as well as the edit distance between two given \strips\ action models.
\end{itemize}

@@

Compiling the learning of action models into classical planning is a general and flexible approach that allows to accommodate various amounts and kinds of input knowledge and opens up a path for addressing further learning and validation tasks. For instance, the example plans in $\Pi$ could be replaced or complemented by a set $\mathcal{O}$ of sequences of observations (i.e., fully or partial state observations with noisy or missing fluents~\cite{SohrabiRU16}), so learning tasks $\Lambda=\tup{\Psi,\Sigma,\mathcal{O},\Xi_0}$ could also be addressed. Furthermore, our approach seems extensible to learning other types of generative models (e.g. hierarchical models like HTN or behaviour trees), that can be more appealing than \strips\ models, since using them to compute planning solutions requires less search effort.

@@

\subsection{Learning from state observations} This learning task corresponds to observing an agent acting in the world but watching only the results of its plan executions, the actual executed actions are unobserved. This learning task is formalized as $\Lambda=\tup{\mathcal{M},\Psi,\mathcal{O}}$:
\begin{itemize}
\item $\mathcal{M}$ is the set of {\em empty} operator schemas, wherein each $\xi\in\mathcal{M}$ is only composed of $head(\xi)$. In some cases, we may not require to start learning from scratch that is, the operator schemas in $\mathcal{M}$ may be not {\em empty} but partially specified operator schemes where some preconditions and effects are a priori known.
\item $\Psi$ is the set of predicates that define the abstract state space of a given planning domain.
\item $\mathcal{O}=\tup{s_0,s_1,\ldots,s_{n}}$ is a sequence of {\em state observations} obtained observing the execution of an {\em unobserved} plan $\pi=\tup{a_1, \ldots, a_n}$ such that, for each {\small $1\leq i\leq n$}, $a_i$ is applicable in $s_{i-1}$ and generates the successor state $s_i=\theta(s_{i-1},a_i)$. Let us also define the learning task  $\Lambda=\tup{\mathcal{M},\Psi,\sigma}$ where the observed states are just the state pair $\sigma=(s_0,s_{n})$, that we call {\em label}, and that comprises the {\em final} state $s_{n}$ resulting from executing the {\em unobserved} plan $\pi$ starting from the {\em initial} state $s_0$. 
\end{itemize}

@@

\subsection{Learning from a labeled plan}
Here we redefine the task of learning \strips\ action models from observations of plan executions to cover the case where the actions executed by the observed agent are known. In this case, the learning task is formalized as $\Lambda=\tup{\mathcal{M},\Psi,\sigma,\pi}$, where:
\begin{itemize}
\item The plan $\pi=\tup{a_1, \ldots, a_n}$, is an action sequence applicable in $s_0\in\sigma$ and that generates the state $s_n\in\sigma$.
\end{itemize}

Figure~\ref{fig:example-lplan} shows an example of a {\em blocksworld} learning task $\Lambda=\tup{\mathcal{M},\Psi,\sigma,\pi}$, that corresponds to observing the execution of an eight-action plan for inverting a four-block tower.

\begin{figure}[hbt!]
{\tt ;;; Predicates in $\Psi$}
\begin{footnotesize}
\begin{verbatim}
(handempty) (holding ?o  - object)
(clear ?o - object) (ontable ?o - object)
(on ?o1 - object ?o2 - object)
\end{verbatim}
\end{footnotesize}

\vspace{0.2cm}

\begin{subfigure}{.25\textwidth}
{\tt ;;; Plan $\pi$}
\begin{footnotesize}
\begin{verbatim}
0: (unstack A B)
1: (putdown A)
2: (unstack B C)
3: (stack B A)
4: (unstack C D)
5: (stack C B)
6: (pickup D)
7: (stack D C)
\end{verbatim}
\end{footnotesize}
\end{subfigure}%
\begin{subfigure}{.6\textwidth}
{\tt ;;; Label $\sigma=(s_0^1,s_{n}^1)$}
\begin{lstlisting}[mathescape]
\end{lstlisting}
\vspace{0.1cm}
\begin{tikzpicture}[node distance = 0mm, block/.style args = {#1,#2}{fill=#1,text width=#2,shape=square}]
\node (initD) [draw]{D};
\node (initC) [draw, above=of initD.north]{C};
\node (initB) [draw, above=of initC.north]{B};
\node (initA) [draw, above=of initB.north]{A};
\draw[thick] (-1,-0.25) -- (2.5,-0.25);

\node (goalA) [draw, right=10mm of initD]{A};
\node (goalB) [draw, right=10mm of initC]{B};
\node (goalC) [draw, right=10mm of initB]{C};
\node (goalD) [draw, right=10mm of initA]{D};
\end{tikzpicture}
\vspace{0.6cm}
\end{subfigure}%
 \caption{\small Example of a $\Lambda=\tup{\mathcal{M},\Psi,\sigma,\pi}$ task for learning a \strips\ action model in the blocksworld from a labeled plan.}
\label{fig:example-lplan}
\end{figure}

\subsection{Learning from multiple plans}
The previous definitions formalize the learning \strips\ action models from observations of a single plan execution. These definitions are extensible to the more general case where learning from the execution of multiple plans:
\begin{itemize}
  \item When learning from states observations, $\Lambda=\tup{\mathcal{M},\Psi,\mathcal{O}}$ is still a valid formalization but considering that now that $\mathcal{O}=\tup{s_0^1,s_1^1,\ldots,s_{n}^1,\ldots,s_0^t,s_1^t,\ldots,s_{n}^t\ldots,s_0^{\tau},s_1^{\tau},\ldots,s_{n}^{\tau}}$ are sequences of {\em state observations} obtained observing the execution of the corresponding {\em unobserved} plan $\pi^t=\tup{a_1^t, \ldots, a_n^t}$, {\tt\small $1\leq t\leq \tau$} such that, for each {\small $1\leq i\leq n$}, $a_i^t$ is applicable in $s_{i-1}^t$ and generates the successor state $s_i^t=\theta(s_{i-1}^t,a_i^t)$.
  \item When learning from a set of labeled plans the task is defined as $\Lambda=\tup{\mathcal{M},\Psi,\Sigma,\Pi}$, where $\Pi=\{\pi_1,\ldots,\pi_{\tau}\}$ is a given set of example plans and $\Sigma=\{\sigma_1,\ldots,\sigma_{\tau}\}$ is the corresponding set of labels such that each plan $\pi_t=\tup{a_1^t, \ldots, a_n^t}$, {\tt\small $1\leq t\leq \tau$}, is an action sequence applicable in $s_0^t\in\sigma_t$ and that generates the state $s_n^t\in\sigma_t$.
\end{itemize}

@@

\item Actions for {\em applying} a programmed operator schema $\xi\in\mathcal{M}$ bound with objects $\omega\subseteq\Omega^{ar(\xi)}$. Given that the operators headers are known, the variables $pars(\xi)$ are bound to the objects in $\omega$ that appear at the same position.
\begin{small}
\begin{align*}
\hspace*{7pt}\pre(\mathsf{apply_{\xi,\omega}})=&\{pre_{f}(\xi)\implies p(\omega)\}_{\forall p\in\Psi,f=p(pars(\xi))}\\
&\cup \{\neg mode_{val}\},\\
\cond(\mathsf{apply_{\xi,\omega}})=&\{del_{f}(\xi)\}\rhd\{\neg p(\omega)\}_{\forall p\in\Psi,f=p(pars(\xi))},\\
&\{add_{f}(\xi)\}\rhd\{p(\omega)\}_{\forall p\in\Psi,f=p(pars(\xi))},\\
&\{mode_{prog}\}\rhd\{\neg mode_{prog}\},\\
&\{\emptyset\}\rhd\{mode_{val}\}.
\end{align*}
\end{small}

@@

@@

To illustrate how our compilation works, Figure~\ref{fig:plan-observations} shows a plan that solves a classical planning task resulting from the compilation. The plan programs and validates the $stack$ schema using the five state observations shown in Figure~\ref{fig:example-observations} as well as previously specified operator schemes for $pickup$, $putdown$ and $unstack$. Note that this compilation is valid for the particular scenario where the observed states are not full states but {\em partial states}, in the sense that some of the observed fluents (either with positive or negative value) are missing. Even more, the compilation is valid for learning when there are missing state observation in $\mathcal{O}$ by removing any reference to the $mode_{val}$ fluent. In that case the classical planner will determine how many {\em apply} actions are necessary between two {\em validate} actions.

@@

\subsection{Learning from a labeled plan}
We extend the compilation to consider labeled plans. Given a learning task $\Lambda=\tup{\mathcal{M},\Psi,\sigma,\pi}$, the compilation outputs a classical planning task $P_{\Lambda'}=\tup{F_{\Lambda'},A_{\Lambda'},I_{\Lambda'},G_{\Lambda'}}$ such that:
\begin{itemize}
\item $F_{\Lambda'}$ extends $F_{\Lambda}$ with $F_{\pi}=\{plan(name(\xi),\Omega^{ar(\xi)},j)\}$, the fluents to code the steps of plan $\pi=\tup{a_1, \ldots, a_n}$. Fluents $at_j$ and $next_{j,j+1}$, {\small $1\leq j< n$}, are also added to represent the current plan step and to iterate through the steps of the plan.
\item $I_{\Lambda'}$ extends $I_{\Lambda}$ with fluents $F_{\pi}$ plus fluents $at_1$ and $\{next_{j,j+1}\}$, {\small $1\leq j<n$}, for indicating the plan step where the action model is validated. Goals are $G_{\Lambda'}=\{test_1\}$, as now the programmed action model is only validated in the single state observation $s_n$, the last state reached by the plan. If more intermediate observed states are known they can however be used for validation adding a $validate_{i}$ action for each observation $s_{i}\in \mathcal{O}$.
\item With respect to $A_{\Lambda'}$.
\begin{enumerate}
\item The actions for {\em programming} the preconditions/effects of a given operator schema $\xi\in\Xi$ and the actions for {\em validating} the programmed action model in the given state observations are the same.
\item The actions for {\em applying} an already programmed operator have an extra precondition $f\in F_{\pi}$ that encodes the current plan step, and extra conditional effects $\{at_{j}\}\rhd\{\neg at_{j},at_{j+1}\}_{\forall j\in [1,n]}$ for advancing to the next plan step. With this mechanism we ensure that these actions are applied in the same order as in the example plan $\pi$.
\end{enumerate}
\end{itemize}

To illustrate this, the classical plan of Figure~\ref{fig:plan-lplan} is a solution to a learning task $\Lambda=\tup{\mathcal{M},\Psi,\sigma,\pi}$ for getting the {\em blocksworld} action model where operator schemes for {\tt\small pickup}, {\tt\small putdown} and {\tt\small unstack} are specified in $\mathcal{M}$. This plan programs and validates the operator schema {\tt\small stack} from {\em blocksworld}, using the plan $\pi$ and label $\sigma$ shown in Figure~\ref{fig:example-lplan}. Plan steps $[0,8]$ program the preconditions of the {\tt\small stack} operator, steps $[9,13]$ program the operator effects and steps $[14,22]$ validate the programmed operators following the plan $\pi$ shown in the Figure~\ref{fig:example-lplan}.

\begin{figure}[hbt!]
{\footnotesize\tt
     {\bf 00} : (program\_pre\_clear\_stack\_v1)\\
     01 : (program\_pre\_handempty\_stack)\\
     02 : (program\_pre\_holding\_stack\_v2)\\
     03 : (program\_pre\_on\_stack\_v1\_v1)\\
     04 : (program\_pre\_on\_stack\_v1\_v2)\\
     05 : (program\_pre\_on\_stack\_v2\_v1)\\
     06 : (program\_pre\_on\_stack\_v2\_v2)\\
     07 : (program\_pre\_ontable\_stack\_v1)\\
     08 : (program\_pre\_ontable\_stack\_v2)\\
     {\bf 09} : (program\_eff\_clear\_stack\_v1)\\
    10 : (program\_eff\_clear\_stack\_v2)\\
    11 : (program\_eff\_handempty\_stack)\\
    12 : (program\_eff\_holding\_stack\_v1)\\
    13 : (program\_eff\_on\_stack\_v1\_v2)\\
    {\bf 14} : (apply\_unstack a b i1 i2)\\
    15 : (apply\_putdown a i2 i3)\\
    16 : (apply\_unstack b c i3 i4)\\
    17 : (apply\_stack b a i4 i5)\\
    18 : (apply\_unstack c d i5 i6)\\
    19 : (apply\_stack c b i6 i7)\\
    20 : (apply\_pickup d i7 i8)\\
    21 : (apply\_stack d c i8 i9)\\
    {\bf 22} : (validate\_1)
}
 \caption{\small Plan for programming and validating the $stack$ schema using plan $\pi$ and label $\sigma$ (shown in Figure~\ref{fig:example-lplan}) as well as previously specified operator schemes for $pickup$, $putdown$ and $unstack$.}
\label{fig:plan-lplan}
\end{figure}

Addressing the learning task $\Lambda=\tup{\mathcal{M},\Psi,\Sigma,\Pi}$ with classical planning requires introducing a small modification to our compilation. In particular, the actions for {\em validating} the labeled plan $\pi_t\in\Pi$ with the corresponding label $\sigma_t\in\Sigma$, {\tt\small $1\leq t\leq \tau$} are now defined as:
\begin{small}
\begin{align*}
\hspace*{7pt}\pre(\mathsf{validate_{t}})=&G_t\cup\{test_j\}_{1\leq j<t}\cup\{\neg test_j\}_{t\leq j\leq \tau}\cup \{\neg mode_{prog}\},\\
\cond(\mathsf{validate_{t}})=&\{\emptyset\}\rhd\{test_t\} \cup \{\neg f\}_{\forall f\in G_t, f \notin I_{t+1}}\cup \{f\}_{\forall f\in I_{t+1}, f \notin G_t}.
\end{align*}
\end{small}

@@

\begin{mylemma}
Soundness. Any classical plan $\pi$ that solves $P_{\Lambda}$ induces an action model $\mathcal{M}'$ that solves $\Lambda=\tup{\mathcal{M},\Psi,\mathcal{O}/\Sigma,\Pi}$.
\end{mylemma}

@@

\begin{mylemma}
Completeness. Any \strips\ action model $\mathcal{M}'$ that solves a $\Lambda=\tup{\mathcal{M},\Psi,\mathcal{O}/\Sigma,\Pi}$ learning task, is computable solving the corresponding classical planning task $P_{\Lambda}$.
\end{mylemma}

\begin{proof}[Proof sketch]
\begin{small}
By definition, $F_v(\xi)\subseteq F_\Lambda$ fully captures the full set of elements that can appear in a \strips\ action schema $\xi\in\mathcal{M}$ given its header and the set of predicates $\Psi$. The compilation does not discard any possible \strips\ action schema definable within $F_v$ that satisfies the state trajectory constraint given by $\mathcal{O}/\Sigma,\Pi$.
\end{small}
\end{proof}

@@

Static predicates can also constrain the space of possible preconditions by looking at the given set of labels $\Sigma$. One can assume that if a precondition $f\in F_v$ (s.t. $f\in F_v$ is an instantiation of a static predicate in $\Omega_v$) is not compliant with the labels in $\Sigma$ then, fluents $pre_f(\xi)$ and actions $\mathsf{programPre_{f,\xi}}$ can be discarded for every $\xi\in\Xi$. For instance in the {\em zenotravel} domain $pre\_next\_board\_v1\_v1$, $pre\_next\_debark\_v1\_v1$, $pre\_next\_fly\_v1\_v1$, $pre\_next\_zoom\_v1\_v1$, $pre\_next\_refuel\_v1\_v1$ can be discarded (and their corresponding programming actions) because a precondition {\tt\small(next ?v1 ?v1 - flevel)} will never hold at any state in $\Sigma$.

@@

\section{Evaluating \strips\ action models}
\label{sec:Section6}
We assess how well a \strips\ action model $\mathcal{M}$ explains a given sequence of observations $\mathcal{O}$ according to the amount of {\em edition} required by $\mathcal{M}$ to induce $\mathcal{O}$. %In the extreme, if $\mathcal{M}$ perfectly explains $\mathcal{O}$, no model {\em edition} is necessary.

\subsection{Edition of \strips\ action models}

We first define the allowed \emph{operations} to edit a given \strips\ action model. With the aim of keeping a tractable branching factor of the planning instances that results from our compilation, we only define two {\em edit operations}:
\begin{itemize}
\item {\em Deletion}. A fluent $pre_f(\xi)/del_f(\xi)/add_f(\xi)$ is removed from the operator schema $\xi\in\mathcal{M}$, $f\in F_v(\xi)$.
\item {\em Insertion}. A fluent $pre_f(\xi)/del_f(\xi)/add_f(\xi)$ is added to the operator schema $\xi\in\mathcal{M}$, $f\in F_v(\xi)$.
\end{itemize}

@@

The $\Lambda$ learning tasks can swap the roles of two operators whose headers match or two action parameters that belong to the same type (e.g. the {\em blocksworld} operator {\small\tt stack} can be {\em learned} with the preconditions and effects of the {\small\tt unstack} operator and vice versa, or the parameters of the {\small\tt stack} operator can be swapped). The {\em observation edit distance} is a semantic measure that is robust to role changes of this kind.

>>>>

A first description of the compilation and its extensions previously appeared in several conference papers~\cite{aineto2018learning}. Compared to the conference papers, the present paper includes the following novel material:
\begin{itemize}
\item Unifies the formulation of our approach for learning \strips\ action models from state observations and from observed plans.
\item Shows that our compilation is also valid for the particular scenario where the observed states are not full states but {\em partial states}, in the sense that some of the observed fluents (either with positive or negative value) are missing.
\item Proposes a distance metric to assess how well a \strips\ action model matches a given set of plans. Defines the compilation to compute this distance as well as the edit distance between two given \strips\ action models.
\end{itemize}

@@

@@

\subsection{Learning from observations of plan executions}

The first learning task corresponds to observing an agent acting in the world but watching only the results of its plan executions, the actual executed actions are unobserved. This learning task is formalized as $\Lambda=\tup{\mathcal{M},\Psi,\mathcal{O}}$:
\begin{itemize}
\item $\mathcal{M}$ is the set of {\em empty} operator schemas, wherein each $\xi\in\mathcal{M}$ is only composed of $head(\xi)$. In some cases, we may not require to start learning from scratch that is, the operator schemas in $\mathcal{M}$ may be not {\em empty} but partially specified operator schemes where some preconditions and effects are a priori known.
\item $\Psi$ is the set of predicates that define the abstract state space of a given planning domain.
\item $\mathcal{O}=\tup{s_0,s_1,\ldots,s_{n}}$ is a sequence of {\em state observations} obtained observing the execution of an {\em unobserved} plan $\pi=\tup{a_1, \ldots, a_n}$ such that, for each {\small $1\leq i\leq n$}, $a_i$ is applicable in $s_{i-1}$ and generates the successor state $s_i=\theta(s_{i-1},a_i)$. The observed states are not required to be full states, they can be {\em partial states}, in the sense that some of the observed fluents (either with positive or negative value) are missing.
\end{itemize}

@@

Here we redefine the task of learning \strips\ action models from observations of plan executions to cover the case where the actions executed by the observed agent are known. In this case there may be missing state observation in $\mathcal{O}$, otherwise the learning task is trivial as previously explained. In this case, the learning task is formalized as $\Lambda=\tup{\mathcal{M},\Psi,\mathcal{O},\pi}$, where:
\begin{itemize}
\item The plan $\pi=\tup{a_1, \ldots, a_n}$, is an action sequence that produces the sequence of state observations given in $\mathcal{O}$.
\end{itemize}

Figure~\ref{fig:example-plans} shows an example of a {\em blocksworld} learning task $\Lambda=\tup{\mathcal{M},\Psi,\mathcal{O},\pi}$, that corresponds to observing the execution of an eight-action plan for inverting a four-block tower. Note that $\mathcal{M},\Psi$ are skipped since they are the same as for Figure~\ref{fig:example-observations}.

\begin{figure}[hbt!]
{\footnotesize\tt ;;;;;; Observations $\mathcal{O}$}
\begin{footnotesize}
\begin{verbatim}
;;; observation #0
(clear B) (on B A) (ontable A) (handempty)

;;; observation #4
(clear A) (on blockA B) (ontable B) (handempty)
\end{verbatim}
\end{footnotesize}

{\footnotesize\tt ;;; Plan $\pi$}
\begin{footnotesize}
\begin{verbatim}
0: (unstack B A)
1: (putdown B)
2: (pickup A)
3: (stack A B)
\end{verbatim}
\end{footnotesize}

 \caption{\small Example of a $\Lambda=\tup{\mathcal{M},\Psi,\mathcal{O},\pi}$ task for learning a \strips\ action model in the blocksworld from a four-action plan and two state observations.}
\label{fig:example-plans}
\end{figure}

The previous definitions formalize the learning \strips\ action models from observations of a single plan execution. These definitions are extensible to the more general case where learning from the execution of multiple plans:
\begin{itemize}
  \item When learning from states observations, $\Lambda=\tup{\mathcal{M},\Psi,\mathcal{O}}$ is still a valid formalization but considering that now that $\mathcal{O}=\tup{s_0^1,s_1^1,\ldots,s_{n}^1,\ldots,s_0^t,s_1^t,\ldots,s_{n}^t\ldots,s_0^{\tau},s_1^{\tau},\ldots,s_{n}^{\tau}}$ are sequences of {\em state observations} obtained observing the execution of the corresponding {\em unobserved} plan $\pi^t=\tup{a_1^t, \ldots, a_n^t}$, {\tt\small $1\leq t\leq \tau$} such that, for each {\small $1\leq i\leq n$}, $a_i^t$ is applicable in $s_{i-1}^t$ and generates the successor state $s_i^t=\theta(s_{i-1}^t,a_i^t)$.
  \item When learning from a set of labeled plans the task is defined as $\Lambda=\tup{\mathcal{M},\Psi,\mathcal{O},\Pi}$, where $\Pi=\{\pi_1,\ldots,\pi_{\tau}\}$ is a given set of example plans producing the given set of state obsevations.
\end{itemize}

@@

\item Actions for {\em applying} a programmed operator schema $\xi\in\mathcal{M}$ bound with objects $\omega\subseteq\Omega^{ar(\xi)}$. Given that the operators headers are known, the variables $pars(\xi)$ are bound to the objects in $\omega$ that appear at the same position. Figure~\ref{fig:compilation} shows the PDDL encoding of the action for applying a programmed operator $stack$.
\begin{small}
\begin{align*}
\hspace*{7pt}\pre(\mathsf{apply_{\xi,\omega}})=&\{pre_{f}(\xi)\implies p(\omega)\}_{\forall p\in\Psi,f=p(pars(\xi))}\\
&\cup \{\neg mode_{val}\},\\
\cond(\mathsf{apply_{\xi,\omega}})=&\{del_{f}(\xi)\}\rhd\{\neg p(\omega)\}_{\forall p\in\Psi,f=p(pars(\xi))},\\
&\{add_{f}(\xi)\}\rhd\{p(\omega)\}_{\forall p\in\Psi,f=p(pars(\xi))},\\
&\{mode_{prog}\}\rhd\{\neg mode_{prog}\},\\
&\{\emptyset\}\rhd\{mode_{val}\}.
\end{align*}
\end{small}

@@

\begin{figure}
\begin{scriptsize}
\begin{verbatim}
(:action apply_stack
  :parameters (?o1 - object ?o2 - object)
  :precondition
   (and (or (not (pre_on_stack_v1_v1)) (on ?o1 ?o1))
        (or (not (pre_on_stack_v1_v2)) (on ?o1 ?o2))
        (or (not (pre_on_stack_v2_v1)) (on ?o2 ?o1))
        (or (not (pre_on_stack_v2_v2)) (on ?o2 ?o2))
        (or (not (pre_ontable_stack_v1)) (ontable ?o1))
        (or (not (pre_ontable_stack_v2)) (ontable ?o2))
        (or (not (pre_clear_stack_v1)) (clear ?o1))
        (or (not (pre_clear_stack_v2)) (clear ?o2))
        (or (not (pre_holding_stack_v1)) (holding ?o1))
        (or (not (pre_holding_stack_v2)) (holding ?o2))
        (or (not (pre_handempty_stack)) (handempty)))
  :effect
   (and (when (del_on_stack_v1_v1) (not (on ?o1 ?o1)))
        (when (del_on_stack_v1_v2) (not (on ?o1 ?o2)))
        (when (del_on_stack_v2_v1) (not (on ?o2 ?o1)))
        (when (del_on_stack_v2_v2) (not (on ?o2 ?o2)))
        (when (del_ontable_stack_v1) (not (ontable ?o1)))
        (when (del_ontable_stack_v2) (not (ontable ?o2)))
        (when (del_clear_stack_v1) (not (clear ?o1)))
        (when (del_clear_stack_v2) (not (clear ?o2)))
        (when (del_holding_stack_v1) (not (holding ?o1)))
        (when (del_holding_stack_v2) (not (holding ?o2)))
        (when (del_handempty_stack) (not (handempty)))
        (when (add_on_stack_v1_v1) (on ?o1 ?o1))
        (when (add_on_stack_v1_v2) (on ?o1 ?o2))
        (when (add_on_stack_v2_v1) (on ?o2 ?o1))
        (when (add_on_stack_v2_v2) (on ?o2 ?o2))
        (when (add_ontable_stack_v1) (ontable ?o1))
        (when (add_ontable_stack_v2) (ontable ?o2))
        (when (add_clear_stack_v1) (clear ?o1))
        (when (add_clear_stack_v2) (clear ?o2))
        (when (add_holding_stack_v1) (holding ?o1))
        (when (add_holding_stack_v2) (holding ?o2))
        (when (add_handempty_stack) (handempty))
        (when (modeProg) (not (modeProg)))))
\end{verbatim}
\end{scriptsize}
 \caption{\small PDDL action for applying an already programmed schema $stack$ (implications coded as disjunctions).}
\label{fig:compilation}
\end{figure}

@@

To illustrate how our compilation works, Figure~\ref{fig:plan-observations} shows a plan that solves a classical planning task resulting from the compilation. The plan programs and validates the $stack$ schema using the five state observations shown in Figure~\ref{fig:example-observations} as well as previously specified operator schemes for $pickup$, $putdown$ and $unstack$. Note that this compilation is valid for the particular scenario where the observed states are not full states but {\em partial states}, in the sense that some of the observed fluents (either with positive or negative value) are missing. Even more, the compilation is valid for learning when there are missing state observations in $\mathcal{O}$ by removing any reference to the $mode_{val}$ fluent. In that case the classical planner will determine how many {\em apply} actions are necessary between two {\em validate} actions.

@@

\subsection{Learning from a plan}
We extend the compilation to consider observed actions. Given a learning task $\Lambda=\tup{\mathcal{M},\Psi,\mathcal{O},\pi}$, the compilation outputs a classical planning task $P_{\Lambda'}=\tup{F_{\Lambda'},A_{\Lambda'},I_{\Lambda'},G_{\Lambda'}}$ such that:
\begin{itemize}
\item $F_{\Lambda'}$ extends $F_{\Lambda}$ with $F_{\pi}=\{plan(name(\xi),\Omega^{ar(\xi)},j)\}$, the fluents to code the steps of plan $\pi=\tup{a_1, \ldots, a_n}$. Fluents $at_j$ and $next_{j,j+1}$, {\small $1\leq j< n$}, are also added to represent the current plan step and to iterate through the steps of the plan.
\item $I_{\Lambda'}$ extends $I_{\Lambda}$ with fluents $F_{\pi}$ plus fluents $at_1$ and $\{next_{j,j+1}\}$, {\small $1\leq j<n$}, for indicating the plan step where the action model is validated. Goals are as in the original compilation, $G_{\Lambda'}=G_{\Lambda}=\bigcup_{1\leq t\leq \tau}\{test_t\}$. In other words, intermediate observed states are used for validation adding a $test_t$ fluent for each observation in $\mathcal{O}$.
\item With respect to $A_{\Lambda'}$.
\begin{enumerate}
\item The actions for {\em programming} the preconditions/effects of a given operator schema $\xi\in\Xi$ and the actions for {\em validating} the programmed action model in the given state observations are the same.
\item The actions for {\em applying} an already programmed operator have an extra precondition $f\in F_{\pi}$ that encodes the current plan step, and extra conditional effects $\{at_{j}\}\rhd\{\neg at_{j},at_{j+1}\}_{\forall j\in [1,n]}$ for advancing to the next plan step. With this mechanism we ensure that these actions are applied in the same order as in the example plan $\pi$.
\end{enumerate}
\end{itemize}

To illustrate this, the classical plan of Figure~\ref{fig:plan-lplan} is a solution to a learning task $\Lambda=\tup{\mathcal{M},\Psi,\mathcal{O},\pi}$ for getting the {\em blocksworld} action model where operator schemes for {\tt\small pickup}, {\tt\small putdown} and {\tt\small unstack} are specified in $\mathcal{M}$. This plan programs and validates the operator schema {\tt\small stack} from {\em blocksworld}, using the plan $\pi$ and state observations $\mathcal{O}$ shown in Figure~\ref{fig:example-plans}. Plan steps $[0,8]$ program the preconditions of the {\tt\small stack} operator, steps $[9,13]$ program the operator effects and steps $[14,18]$ validate the programmed operators following the plan $\pi$ shown in the Figure~\ref{fig:example-plans}.

\begin{figure}[hbt!]
{\footnotesize\tt
     {\bf 00} : (program\_pre\_clear\_stack\_v1)\\
     01 : (program\_pre\_handempty\_stack)\\
     02 : (program\_pre\_holding\_stack\_v2)\\
     03 : (program\_pre\_on\_stack\_v1\_v1)\\
     04 : (program\_pre\_on\_stack\_v1\_v2)\\
     05 : (program\_pre\_on\_stack\_v2\_v1)\\
     06 : (program\_pre\_on\_stack\_v2\_v2)\\
     07 : (program\_pre\_ontable\_stack\_v1)\\
     08 : (program\_pre\_ontable\_stack\_v2)\\
     {\bf 09} : (program\_eff\_clear\_stack\_v1)\\
    10 : (program\_eff\_clear\_stack\_v2)\\
    11 : (program\_eff\_handempty\_stack)\\
    12 : (program\_eff\_holding\_stack\_v1)\\
    13 : (program\_eff\_on\_stack\_v1\_v2)\\
    {\bf 14} : (apply\_unstack b a i1 i2)\\
    15 : (apply\_putdown b i2 i3)\\
    16 : (apply\_pickup a i3 i4)\\
    17 : (apply\_stack a b i4 i5)\\
    {\bf 18} : (validate\_1)
}
 \caption{\small Plan for programming and validating the $stack$ schema using plan $\pi$ and state observations $\mathcal{O}$ (shown in Figure~\ref{fig:example-plans}) as well as previously specified operator schemes for $pickup$, $putdown$ and $unstack$.}
\label{fig:plan-lplan}
\end{figure}

Addressing the learning task $\Lambda=\tup{\mathcal{M},\Psi,\mathcal{O},\Pi}$ with classical planning requires introducing a small modification to our compilation. In particular, the actions for {\em validating} the plan $\pi_t\in\Pi$ in the corresponding sequence of state observations $\tup{s_0^t, s_1^t, \ldots, s_n^t}$, {\tt\small $1\leq t\leq \tau$}, are now defined as:
\begin{small}
\begin{align*}
\hspace*{7pt}\pre(\mathsf{validate_{t}})=&G_t\cup\{test_j\}_{1\leq j<t}\cup\{\neg test_j\}_{t\leq j\leq \tau}\cup \{\neg mode_{prog}\},\\
\cond(\mathsf{validate_{t}})=&\{\emptyset\}\rhd\{test_t\} \cup \{\neg f\}_{\forall f\in G_t, f \notin I_{t+1}}\cup \{f\}_{\forall f\in I_{t+1}, f \notin G_t}.
\end{align*}
\end{small}

The compilation could be extended for learning when there are missing actions in $\pi$ by not coding the corresponding $plan(name(\xi),\Omega^{ar(\xi)},j$ fluent and not adding the extra precondition $f\in F_{\pi}$ to the corresponding corresponding {\em apply} action. In that case the classical planner will determine the {\em apply} action necessary for the validating the learned action model.

@@

\begin{mylemma}
Soundness. Any classical plan $\pi$ that solves $P_{\Lambda}$ induces an action model $\mathcal{M}'$ that solves $\Lambda=\tup{\mathcal{M},\Psi,\mathcal{O},\Pi}$.
\end{mylemma}

@@

\begin{mylemma}
Completeness. Any \strips\ action model $\mathcal{M}'$ that solves a $\Lambda=\tup{\mathcal{M},\Psi,\mathcal{O},\Pi}$ learning task, is computable solving the corresponding classical planning task $P_{\Lambda}$.
\end{mylemma}

\begin{proof}[Proof sketch]
\begin{small}
By definition, $F_v(\xi)\subseteq F_\Lambda$ fully captures the full set of elements that can appear in a \strips\ action schema $\xi\in\mathcal{M}$ given its header and the set of predicates $\Psi$. The compilation does not discard any possible \strips\ action schema definable within $F_v$ that satisfies the state trajectory constraint given by $\mathcal{O},\Pi$.
\end{small}
\end{proof}

@@

Static predicates can also constrain the space of possible preconditions by looking at the given set of state observation $\mathcal{O}$. One can assume that if a precondition $f\in F_v$ (s.t. $f\in F_v$ is an instantiation of a static predicate in $\Omega_v$) is not compliant with the observations in $\mathcal{O}$ then, fluents $pre_f(\xi)$ and actions $\mathsf{programPre_{f,\xi}}$ can be discarded for every $\xi\in\mathcal{M}$. For instance in the {\em zenotravel} domain $pre\_next\_board\_v1\_v1$, $pre\_next\_debark\_v1\_v1$, $pre\_next\_fly\_v1\_v1$, $pre\_next\_zoom\_v1\_v1$, $pre\_next\_refuel\_v1\_v1$ can be discarded (and their corresponding programming actions) because a precondition {\tt\small(next ?v1 ?v1 - flevel)} will never hold at any state in $\mathcal{O}$.

@@

\section{Evaluating \strips\ action models}
\label{sec:Section6}
We assess how well a \strips\ action model $\mathcal{M}$ explains given observations of plan executions according to the amount of {\em edition} required by $\mathcal{M}$ to induce that observations. %In the extreme, if $\mathcal{M}$ perfectly explains $\mathcal{O}$, no model {\em edition} is necessary.

We first define the allowed \emph{operations} to edit a given \strips\ action model. With the aim of keeping a tractable branching factor of the planning instances that will results from our compilation, we only define two {\em edit operations}:
\begin{itemize}
\item {\em Deletion}. A fluent $pre_f(\xi)/del_f(\xi)/add_f(\xi)$ is removed from the operator schema $\xi\in\mathcal{M}$, $f\in F_v(\xi)$.
\item {\em Insertion}. A fluent $pre_f(\xi)/del_f(\xi)/add_f(\xi)$ is added to the operator schema $\xi\in\mathcal{M}$, $f\in F_v(\xi)$.
\end{itemize}

@@

This edit distance can also be defined with respect to a set of plans.

\begin{mydefinition}
  Given an action model $\mathcal{M}$, built from $F_v$ and a set of plans $\Pi=\{\pi_1,\ldots,\pi_{\tau}\}$. The {\bf plans edit distance}, denoted by  $\delta(\mathcal{M},\Pi)$, is the minimal edit distance from $\mathcal{M}$ to any model $\mathcal{M}'$, defined also within $F_v$, such that $\mathcal{M}'$ can produce all the plans $\pi_t\in \Pi$, {\tt\small $1\leq t\leq \tau$}; \[\delta(\mathcal{M},\Pi)=\min_{\forall \mathcal{M}' \rightarrow \Pi} \delta(\mathcal{M},\mathcal{M}')\]
\end{mydefinition}
